---
title: "Tuna Fisheries in Kiribati"
subtitle: "A data-driven overview based on FAO Global Capture Production"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    fig-dpi: 150
  docx:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---
```{r setup}
#| include: false
library(ggplot2)
library(scales)
library(dplyr)
library(sf)
library(mregions2)
library(rnaturalearth)
library(rnaturalearthdata)

source("R/colours.R")
source("R/theme.R")
source("R/data.R")
```

## Introduction

Kiribati is one of the most tuna-dependent nations in the world. Spanning
3.5 million km2 of Pacific Ocean across three island groups (the Gilbert,
Phoenix, and Line Islands), its Exclusive Economic Zone (EEZ) is among the largest
on Earth and sits at the heart of the world's most productive skipjack tuna
grounds. This report examines historical and recent trends in tuna catches
reported to the FAO, broken down by species.

> Data covers FAO-reported national catch statistics and may not reflect the
> full extent of fishing activity within Kiribati's EEZ, including licensed
> foreign vessel catches attributed to other flags.

## Historical catch trends

### Total tuna catch over time

```{r fig-total-trend}
#| fig-cap: "Total tuna catch reported by Kiribati, all species combined"
#| fig-height: 4

kir |>
  group_by(PERIOD) |>
  summarise(total = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = PERIOD, y = total)) +
  geom_area(fill = clr$ocean_navy, alpha = 0.15) +
  geom_line(colour = clr$ocean_navy, linewidth = 0.9) +
  scale_y_continuous(labels = label_comma(scale = 1e-3, suffix = "k t")) +
  scale_x_continuous(breaks = seq(1950, 2025, 10)) +
  labs(
    title    = "Total tuna catch in Kiribati",
    subtitle = "All species, FAO-reported landings",
    x = NULL, y = "Catch (thousand tonnes)",
    caption  = caption_fao
  ) +
  theme_tunaki()
```

### Catch by species over time

```{r fig-species-trend}
#| fig-cap: "Catch by tuna species in Kiribati"
#| fig-height: 5

kir |>
  group_by(PERIOD, Sp_Name_En) |>
  summarise(catch = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = PERIOD, y = catch,
             colour = Sp_Name_En, fill = Sp_Name_En)) +
  geom_area(alpha = 0.12, position = "identity") +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = species_colours) +
  scale_fill_manual(values = species_colours) +
  scale_y_continuous(labels = label_comma(scale = 1e-3, suffix = "k t")) +
  scale_x_continuous(breaks = seq(1950, 2025, 10)) +
  labs(
    title    = "Tuna catch by species in Kiribati",
    subtitle = "FAO-reported landings",
    x = NULL, y = "Catch (thousand tonnes)",
    caption  = caption_fao
  ) +
  theme_tunaki()
```

### Stacked species composition

```{r fig-stacked}
#| fig-cap: "Stacked species composition of tuna catch in Kiribati"
#| fig-height: 4

kir |>
  group_by(PERIOD, Sp_Name_En) |>
  summarise(catch = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = PERIOD, y = catch, fill = Sp_Name_En)) +
  geom_area(position = "stack", alpha = 0.85) +
  scale_fill_manual(values = species_colours) +
  scale_y_continuous(labels = label_comma(scale = 1e-3, suffix = "k t")) +
  scale_x_continuous(breaks = seq(1950, 2025, 10)) +
  labs(
    title    = "Species composition of tuna catch in Kiribati",
    subtitle = "Stacked total, FAO-reported landings",
    x = NULL, y = "Catch (thousand tonnes)",
    caption  = caption_fao
  ) +
  theme_tunaki()
```

### Decadal averages

```{r fig-decadal}
#| fig-cap: "Mean annual catch per decade by species, Kiribati"
#| fig-height: 4

kir |>
  mutate(decade = paste0(floor(PERIOD / 10) * 10, "s")) |>
  group_by(decade, Sp_Name_En) |>
  summarise(mean_catch = mean(VALUE, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = decade, y = mean_catch, fill = Sp_Name_En)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = species_colours) +
  scale_y_continuous(labels = label_comma(scale = 1e-3, suffix = "k t")) +
  labs(
    title    = "Mean annual tuna catch per decade in Kiribati",
    subtitle = "By species",
    x = NULL, y = "Mean catch (thousand tonnes)",
    caption  = caption_fao
  ) +
  theme_tunaki()
```

## Recent snapshot (`r latest_year`)

### Species breakdown

```{r fig-latest-bar}
#| fig-cap: !expr paste("Tuna catch by species in Kiribati,", latest_year)
#| fig-height: 3.5

kir |>
  filter(PERIOD == latest_year) |>
  group_by(Sp_Name_En) |>
  summarise(catch = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  mutate(Sp_Name_En = reorder(Sp_Name_En, catch)) |>
  ggplot(aes(x = catch, y = Sp_Name_En, fill = Sp_Name_En)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(
    aes(label = label_comma(scale = 1e-3, suffix = "k t")(catch)),
    hjust = -0.15, size = 3, colour = clr$basalt, family = "roboto"
  ) +
  scale_fill_manual(values = species_colours) +
  scale_x_continuous(
    labels = label_comma(scale = 1e-3, suffix = "k t"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title    = paste("Tuna catch by species in Kiribati,", latest_year),
    subtitle = "FAO-reported landings",
    x = NULL, y = NULL,
    caption  = caption_fao
  ) +
  theme_tunaki() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = clr$arctic_white)
  )
```

### 10-year trend

```{r fig-recent-trend}
#| fig-cap: !expr paste("Tuna catch by species, Kiribati,", latest_year - 9, "to", latest_year)
#| fig-height: 4

kir |>
  filter(PERIOD >= latest_year - 9) |>
  group_by(PERIOD, Sp_Name_En) |>
  summarise(catch = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = PERIOD, y = catch, colour = Sp_Name_En)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = species_colours) +
  scale_y_continuous(labels = label_comma(scale = 1e-3, suffix = "k t")) +
  scale_x_continuous(breaks = seq(latest_year - 9, latest_year, 2)) +
  labs(
    title    = "Recent tuna catch trends in Kiribati",
    subtitle = paste(latest_year - 9, "to", latest_year),
    x = NULL, y = "Catch (thousand tonnes)",
    caption  = caption_fao
  ) +
  theme_tunaki()
```

### Kiribati in Pacific context

```{r fig-pacific-context}
#| fig-cap: !expr paste("Top 15 Pacific tuna fishing nations,", latest_year)
#| fig-height: 4.5

data |>
  filter(
    Sp_Name_En %in% tuna_species,
    Ocean_Group_En == "Pacific Ocean",
    PERIOD == latest_year
  ) |>
  group_by(Country_Name_En) |>
  summarise(catch = sum(VALUE, na.rm = TRUE), .groups = "drop") |>
  slice_max(catch, n = 15) |>
  mutate(
    Country_Name_En = reorder(Country_Name_En, catch),
    highlight = Country_Name_En == "Kiribati"
  ) |>
  ggplot(aes(x = catch, y = Country_Name_En, fill = highlight)) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c("FALSE" = clr$glacial_blue, "TRUE" = clr$lobster_orange),
    guide  = "none"
  ) +
  scale_x_continuous(
    labels = label_comma(scale = 1e-3, suffix = "k t"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = paste("Top 15 Pacific tuna fishing nations,", latest_year),
    subtitle = "Kiribati highlighted in orange",
    x = NULL, y = NULL,
    caption  = caption_fao
  ) +
  theme_tunaki() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = clr$arctic_white)
  )
```

## Summary

```{r summary-table}
#| tbl-cap: "Summary statistics for Kiribati tuna catch by species"

kir |>
  group_by(Species = Sp_Name_En) |>
  summarise(
    `First year`     = min(PERIOD[VALUE > 0], na.rm = TRUE),
    `Peak year`      = PERIOD[which.max(VALUE)],
    `Peak catch (t)` = label_comma()(max(VALUE, na.rm = TRUE)),
    `Latest (t)`     = label_comma()(sum(VALUE[PERIOD == latest_year],
                                         na.rm = TRUE)),
    .groups = "drop"
  ) |>
  knitr::kable()
```